---
title: "Calculating Pre/Post Correlation from the Standard Deviation of Change Scores"
author: "Matthew B. Jané"
date: 2023-09-08
description: "Welcome to the first blog post of *Meta-Analysis Tricks*! In this post we will go over a couple of ways to directly calculate pre/post correlations from alternative statistics such as standard deviations of change scores. Pre/post correlations are necessary to obtain various types of effect sizes and standard errors such as the repeated measures variant of Cohen's $d$, also referred to as Cohen's $d_{rm}$."
categories:
  - repeated measures
  - correlations

title-meta: "Calculating Pre/Post Correlation from the Standard Deviation of Change Scores"
author-meta: "Matthew B. Jané"
date-meta: 2023-09-08
---

## Step 1: Obtain Pre-test, Post-test, and Change score standard deviations

In order to calculate the pre/post correlation ($r$), we need an estimate of the standard deviation (SD) of pre-test scores ($S_{pre}$), the SD of post-test scores ($S_{post}$), and the SD of change scores ($S_{change}$, where $x_{change}=x_{pre}-x_{post}$). If the post-test standard deviation is unavailable, but the pre-test standard deviation is available, you can approximate the post-test standard deviation by taking average ratio of the pre-test SD and post-test SD from all other studies in the current meta-analysis.
 
$$
\bar{S}_{ratio}=\frac{1}{k}\sum_{i=1}^k \frac{S_{post,i}}{S_{pre,i}}
$$
Where $k$ is the number of studies in the meta-analysis. Then we can make an approximation of the post-test standard by multiplying the pre-test SD by the ratio of SDs,

$$
S_{post}\approx \bar{S}_{ratio}\times S_{pre}
$$

A rougher approximation would be to simply set the pre-test SD and post-test SD to be equal. 


```{r}
# Define standard deviations
S_pre = 9
S_post = 11
S_change = 8
```

## Step 2: Calculate the Pre/Post Correlation

The correlation between pre-test and post-test scores ($r$) can be calculated by re-arranging the equation for change score SD:

$$
S_{change} = \sqrt{S^2_{pre} + S^2_{post} - 2rS_{pre}S_{post}}
$$
Then we can solve for $r$,

$$
r = \frac{S^2_{pre} + S^2_{post} - S^2_{change}}{2S_{pre}S_{post}}
$$
Note that this is a direct conversion and not merely an approximation.

```{r}
# Calculate pre/post correlation
r = (S_pre^2 + S_post^2 - S_change^2) / (2*S_pre*S_post)

# Print results
print(paste0('r = ',round(r,3)))
```


## Applying it to a simulated dataset

We can simulate correlated pre/post scores from a bivariate Gaussian with known parameters. It can be seen that the correlation calculated from the formulas above is perfectly precise.

```{r}
# install.packages('MASS')
library(MASS)

# Define parameters
S_pre = 9
S_post = 11
r_true = .70

# Simulate correlated pre/post scores from bivariate gaussian
data = mvrnorm(n=200,
               mu=c(0,0),
               Sigma = data.frame(x=c(S_pre^2,r_true*S_pre*S_post),
                                  y=c(r_true*S_pre*S_post,S_post^2)),
               empirical = TRUE)

# Obtain simulated scores
x_pre = data[,1] # Pre-test scores
x_post = data[,2] # Post-test scores
x_change = x_post - x_pre # Calculate change scores

# Calculate standard deviations
S_pre = sd(x_pre)
S_post = sd(x_post)
S_change = sd(x_change)

# Calculate pre/post correlation
r = (S_pre^2 + S_post^2 - S_change^2) / (2*S_pre*S_post)

print(paste0('r = ',r))
```







